QuickSched: Task-based parallelism with dependencies and conflicts

### 2014-09-29 paper/paper.tex
fixed description of BH solver, need to add correct numbers and figures.

These tasks can be created recursively over the cell hierarchy
as shown in the function {\tt make\_tasks} in \fig{MakeTasks}.
The function is called on the root cell with the root cell
and {\tt NULL} as its two cell parameters.
The function recurses as follows:
\begin{itemize}
    \item If called with a single (line~6), split (line~7) cell,
        recurse over all the cell's sub-cells (line~9), and all
        pairs of the cell's sub-cells (line~11),
    \item If called with a single unsplit cell (line~13),
        create a self-interaction task on that cell (line~14),
    \item If called with two cells that are sufficiently well
        separated (line~21), create two particle-cell pair
        interactions (lines~23 and~28) over both cells in
        opposite orders, which depend on the center of mass
        task of each cell,
    \item If called with two cells that are not well
        separated and both cells are split (line~33),
        recurse over all pairs of sub-cells spanning
        both cells (line~37), and
    \item If called with two cells that are not well separated
        and either of the cells are not split, create
        a particle-particle pair task over both cells.
\end{itemize}
\noindent where every interaction task additionally locks
the cells on which it operates (lines~16, 25, 30, and 42--43).

In order to prevent generating
a large number of very small tasks, the task generation only recurses
if the cells contain more than a minimum number $n_\mathsf{task}$
of threads each (lines~7 and~34).
The tasks themselves are then left to recurse over the sub-trees,
which is why in these cases, the tasks are made to depend on the
center of mass tasks (lines~17--18 and~41--47)
which may be used in the ensuing interactions.

This Barnes-Hut tree-code was used to approximate the gravitational
N-Body problem for 1\,000\,000 particles with random coordinates
in $[0,1]^3$.
The parameters $n_\mathsf{max}=100$ and $n_\mathsf{task}=5000$
were used to generate the tasks, and cell pairs were considered
well separated if not directly adjacent.
Using the above scheme generated 97\,553 tasks, of which
512 self-interaction tasks, ??? particle-particle interaction
task, ??? particle-cell interaction tasks, and 37\,449
center of mass tasks.
A total of 141\,840 dependencies were generated, along with
104\,392 locks on 37\,449 resources.

For these tests, OpenMP parallelism was used and resource
re-owning was switched off.
Resource ownership was attributed by dividing the global
{\tt parts} array by the number of queues and assigning each cell's
{\tt res} to the fraction of the {\tt parts} array to which
the first of its own {\tt parts} belong.
The interactions were computed 10 times for each number of
cores, and the average thereof taken for the scaling and
efficiency results in \fig{BHResults}.
The timings are for {\tt qsched\_run}, including the cost of
{\tt qsched\_start}, which does not run in parallel.
Setting up the scheduler, tasks, and resources took, in all
cases, an average of 51.3\,ms.

@@

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}[basicstyle=\scriptsize\tt]
void comp_com(struct cell *c) {
  int j, k;
  c->com[0] = 0.0; c->com[1] = 0.0; c->com[2] = 0.0;
  c->mass = 0.0;
  if (c->split)
    for (k = 0; k < 8; k++) {
      struct cell *cp = c->progeny[k];
      for (j = 0; j < 3; j++)
        c->com[j] += cp->com[j] * cp->mass;
      c->mass += cp->mass;
    }
  else
    for (k = 0; k < 8; k++) {
      struct part *p = &c->parts[k];
      for (j = 0; j < 3; j++)
        c->com[j] += p->x[j] * p->mass;
      c->mass += p->mass;
    }
  c->com[0] /= c->mass; c->com[1] /= c->mass; c->com[2] /= c->mass;
}

void comp_self(struct cell *c) {
  int j, k;
  if (c->split)
    for (j = 0; j < 8; j++) {
      comp_self(c->progeny[j]);
      for (k = j + 1; k < 8; k++)
        comp_pair(c->progeny[j], c->progeny[k]);
    }
  else
    for (j = 0; j < c->count; j++)
      for (k = j + 1; k < c->count; k++)
        interact c->parts[j] and c->parts[k].
}

void comp_pair(struct cell *ci, struct cell *cj) {
  int j, k;
  if (ci and cj well separated) {
    comp_pair_pc(ci, cj);
    comp_pair_pc(cj, ci);
  } else if (ci->split && cj->split)
    for (j = 0; j < 8; j++)
      for (k = 0; k < 8; k++)
        comp_pair(ci->progeny[j], cj->progeny[k]);
  else
    for (j = 0; j < ci->count; j++)
      for (k = 0; k < cj->count; k++)
        interact ci->parts[j] and cj->parts[k].
}

void comp_pair_cp(struct cell *ci, struct cell *cj) {
  int k;
  for (k = 0; k < ci->count; k++)
    interact ci->parts[k] and cj center of mass.
}
    \end{lstlisting}
\end{minipage}\end{center}
    \caption{Task functions for the Barnes-Hut tree-code.}
    \label{fig:BHTasks}
\end{figure}

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}[basicstyle=\scriptsize\tt]
enum { tSELF , tPAIR_PP , tPAIR_PC , tCOM };
void make_tasks(struct qsched *s, struct cell *ci, struct cell *cj) {
  int j, k;
  qsched_task_t tid;
  struct cell *data[2];
  if (cj == NULL) {
    if (ci->split && ci->count > n_task)
      for (j = 0; j < 8; j++) {
        make_tasks(s, ci->progeny[j], NULL);
        for (k = j + 1; k < 8; k++)
          make_tasks(s, ci->progeny[j], ci->progeny[k]);
      }
    else {
      tid = qsched_addtask(s, tSELF, qsched_flags_none, &ci,
                           sizeof(struct cell *), ci->count * ci->count);
      qsched_addlock(s, tid, ci->res);
      if (ci->split) qsched_addunlock(s, ci->com, tid);
    }
  } else if (ci and cj are well separated) {
    data[0] = ci; data[1] = cj;
    tid = qsched_addtask(s, tPAIR_PC, qsched_flags_none, data,
                         sizeof(struct cell *) * 2, ci->count);
    qsched_addlock(s, tid, ci->res);
    qsched_addunlock(s, cj->com, tid);
    data[0] = cj; data[1] = ci;
    tid = qsched_addtask(s, tPAIR_PC, qsched_flags_none, data,
                         sizeof(struct cell *) * 2, cj->count);
    qsched_addlock(s, tid, cj->res);
    qsched_addunlock(s, ci->com, tid);
  } else if (ci->split && cj->split && ci->count * cj->count > n_task * n_task)
    for (j = 0; j < 8; j++)
      for (k = 0; k < 8; k++) make_tasks(s, ci->progeny[j], cj->progeny[k]);
  else {
    data[0] = ci; data[1] = cj;
    tid = qsched_addtask(s, tPAIR_PP, qsched_flags_none, data,
                         sizeof(struct cell *) * 2, ci->count * cj->count);
    qsched_addlock(s, tid, ci->res);
    qsched_addlock(s, tid, cj->res);
    if (ci->split && cj->split) {
      qsched_addunlock(s, ci->com, tid);
      qsched_addunlock(s, cj->com, tid);
    }
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
    \caption{C-like pseudo-code for recursive task creation
        for the Barnes-Hut tree-code.}
    \label{fig:MakeTasks}
\end{figure}

The functions for the task themselves are relatively
straight-forward and shown in \fig{BHTasks}, and the
execution function can be written as:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void exec_fun ( int type , void *data ) {
  struct cell **cells = (struct cell **)data;
  switch ( type ) {
    case tSELF:
      comp_self( cells[0] );
      break;
    case tPAIR_PP:
      comp_pair( cells[0] , cells[1] );
      break;
    case tPAIR_PC:
      comp_pair_pc( cells[0] , cells[1] );
      break;
    case tCOM:
      comp_com( cells[0] );
      break;
    default:
      error( "Unknown task type." );
    }
  }
    \end{lstlisting}
\end{minipage}\end{center}

>>>>

These tasks can be created recursively over the cell hierarchy
as shown in the function {\tt make\_tasks} in \fig{MakeTasks}.
The function is called on the root cell with the root cell
and {\tt NULL} as its two cell parameters.
The function recurses as follows:
\begin{itemize}
    \item If called with a single (line~6), split (line~7) cell,
        recurse over all the cell's sub-cells (line~9), and all
        pairs of the cell's sub-cells (line~11),
    \item If called with a single unsplit cell (line~13),
        create a self-interaction task as well as a particle-cell
        task on that cell (line~14),
    \item If called with two non-neighbouring cells (line~21),
        do nothing, as these interactions
        will be computed by the particle-cell task,
    \item If called with two neighbouring cells and both cells
        are split (line~33),
        recurse over all pairs of sub-cells spanning
        both cells (line~37), and
    \item If called with two neighbouring cells
        and one of the cells are not split, create
        a particle-particle pair task over both cells.
\end{itemize}
\noindent where every interaction task additionally locks
the cells on which it operates (lines~16, 25, 30, and 42--43).
In order to prevent generating
a large number of very small tasks, the task generation only recurses
if the cells contain more than a minimum number $n_\mathsf{task}$
of threads each (lines~7 and~34).

The particle-particle pair interaction tasks are implemented
by computing the interactions between all particle pairs spanning
both cells in a double for-loop.
The particle-cell interactions for each leaf node are computed by
traversing the tree recursively starting from the root node:
\begin{itemize}
  \item If called with a node that is a hierarchical parent of
    the leaf node, or with a node that is a direct neighbour of
    a hierarchical parent of the leaf node, recurse over the
    node's sub-cells,
  \item Otherwise, compute the interaction between the leaf node's
    center of mass and all the particles in the leaf node.
\end{itemize}

This task decomposition differs from the traditional tree-walk
in the Barnes-Hut algorithm in that the particle-cell interactions
are grouped per leaf, with each leaf doing its own tree walk.
This approach was chosen to maximize the memory locality
of the particle-cell calculation, as the particles in the leaf,
which are traversed for each particle-cell interaction, are
contiguous in memory, and are thus more likely to remain in the
lowest-level cache.

This Barnes-Hut tree-code was used to approximate the gravitational
N-Body problem for 1\,000\,000 particles with uniformly random coordinates
in $[0,1]^3$.
The parameters $n_\mathsf{max}=100$ and $n_\mathsf{task}=5000$
were used to generate the tasks.
Using the above scheme generated 97\,553 tasks, of which
512 self-interaction tasks, ??? particle-particle interaction
task, and ??? particle-cell interaction tasks.
A total of 141\,840 dependencies were generated, along with
104\,392 locks on 37\,449 resources.

For these tests, {\tt pthread}s parallelism was used and resource
re-owning was switched off.
Resource ownership was attributed by dividing the global
{\tt parts} array by the number of queues and assigning each cell's
{\tt res} to the fraction of the {\tt parts} array to which
the first of its own {\tt parts} belong.
The interactions were computed 10 times for each number of
cores, and the average thereof taken for the scaling and
efficiency results in \fig{BHResults}.
The timings are for {\tt qsched\_run}, including the cost of
{\tt qsched\_start}, which does not run in parallel.
Setting up the scheduler, tasks, and resources took, in all
cases, an average of 51.3\,ms.

@@

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}[basicstyle=\scriptsize\tt]
void comp_self(struct cell *c) {
  if (c->split)
    for (int j = 0; j < 8; j++) {
      comp_self(c->progeny[j]);
      for (int k = j + 1; k < 8; k++)
        comp_pair(c->progeny[j], c->progeny[k]);
    }
  else
    for (int j = 0; j < c->count; j++)
      for (int k = j + 1; k < c->count; k++)
        interact c->parts[j] and c->parts[k].
}

void comp_pair(struct cell *ci, struct cell *cj) {
  if (ci and cj are not neighbours)
    return;
  if (ci->split && cj->split) {
    for (int j = 0; j < 8; j++)
      for (int k = 0; k < 8; k++)
        comp_pair(ci->progeny[j], cj->progeny[k]);
  } else {
    for (int j = 0; j < ci->count; j++)
      for (int k = 0; k < cj->count; k++)
        interact ci->parts[j] and cj->parts[k].
  }
}

void comp_pair_cp(struct cell *leaf, struct cell *c) {
  if (c is a parent of leaf ||
      c is a neighbour of a parent of leaf) {
    for (int k = 0; k < 8; k++)
     comp_pair_cp(leaf, c->progeny[k]);
  } else if (leaf and c are not direct neighbours) {
    for (int k = 0; k < leaf->count; k++)
      interact leaf->parts[k] and c center of mass.
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
    \caption{Task functions for the Barnes-Hut tree-code.}
    \label{fig:BHTasks}
\end{figure}

\begin{figure}
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}[basicstyle=\scriptsize\tt]
enum { tSELF , tPAIR_PP , tPAIR_PC };
void make_tasks(struct qsched *s, struct cell *ci, struct cell *cj) {
  int j, k;
  qsched_task_t tid;
  struct cell *data[2];
  if (cj == NULL) {
    if (ci->split && ci->count > n_task)
      for (j = 0; j < 8; j++) {
        make_tasks(s, ci->progeny[j], NULL);
        for (k = j + 1; k < 8; k++)
          make_tasks(s, ci->progeny[j], ci->progeny[k]);
      }
    else {
      tid = qsched_addtask(s, tSELF, qsched_flags_none, &ci,
                           sizeof(struct cell *),
                           ci->count * ci->count);
      qsched_addlock(s, tid, ci->res);
      tid = qsched_addtask(s, tPAIR_PC, qsched_flags_none, &ci,
                           sizeof(struct cell *), ci->count);
      qsched_addlock(s, tid, ci->res);
    }
  } else if (ci->split && cj->split &&
             ci->count * cj->count > n_task * n_task)
    for (j = 0; j < 8; j++)
      for (k = 0; k < 8; k++)
        make_tasks(s, ci->progeny[j], cj->progeny[k]);
  else {
    data[0] = ci; data[1] = cj;
    tid = qsched_addtask(s, tPAIR_PP, qsched_flags_none, data,
                         sizeof(struct cell *) * 2,
                         ci->count * cj->count);
    qsched_addlock(s, tid, ci->res);
    qsched_addlock(s, tid, cj->res);
    if (ci->split && cj->split) {
      qsched_addunlock(s, ci->com, tid);
      qsched_addunlock(s, cj->com, tid);
    }
  }
}
    \end{lstlisting}
\end{minipage}\end{center}
    \caption{C-like pseudo-code for recursive task creation
        for the Barnes-Hut tree-code.}
    \label{fig:MakeTasks}
\end{figure}

The functions for the task themselves are relatively
straight-forward and shown in \fig{BHTasks}, and the
execution function can be written as:
\begin{center}\begin{minipage}{0.9\textwidth}
    \begin{lstlisting}
void exec_fun ( int type , void *data ) {
  struct cell **cells = (struct cell **)data;
  switch ( type ) {
    case tSELF:
      comp_self( cells[0] );
      break;
    case tPAIR_PP:
      comp_pair( cells[0] , cells[1] );
      break;
    case tPAIR_PC:
      comp_pair_pc( cells[0] , root );
      break;
    case tCOM:
      comp_com( cells[0] );
      break;
    default:
      error( "Unknown task type." );
    }
  }
    \end{lstlisting}
\end{minipage}\end{center}
